Embedding Intuitionistic into Classical Logic

### 2022-10-05 FoSSaCS/fossacs-EIICL.tex
Format change for FoSSaCS

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles.
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

\usepackage{wrapfig}
\usepackage{tikz-qtree}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{arrows,shapes,quotes}

\usepackage{todonotes}

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Embedding Intuitionistic into Classical Logic} %TODO Please add

%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

\author{Alexander {Pluska}}{Faculty of Mathematics, Universtät Wien, Austria}{e11941874@student.tuwien.ac.at}{https://orcid.org/my-orcid?orcid=0000-0002-7709-3335}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Florian Zuleger}{Institute of Logic and Computation, Technische Universität Wien, Austria}{florian.zuleger@tuwien.ac.at}{https://orcid.org/0000-0003-1468-8398}{}

\authorrunning{A. Pluska and F. Zuleger} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Alexander Pluska and Florian Zuleger} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{\textcolor{red}{Theory of Computation Logic Constructive Mathematics}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Intuitionistic Logic, Automated Reasoning, Propositional Logic, First-order Logic, QBF} %TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

%TODO mandatory: add short abstract of the document
\begin{abstract}
The famous double negation translation~\cite{glivenko1929quelques, godel1933intuitionistischen} establishes an embedding from classical into intuitionistic logic. Curiously the reverse direction has not been covered in literature. We present an effective embedding from intuitionistic into classical logic, both in the propositional and first-order case, as well as an effective embedding of intuitionistic propositional logic into quantified boolean formulas.
Furthermore, in the propositional case we establish a parameter that bounds the proof complexity of intuitionistic formulas. As part of our construction, we also present a Tseytin-like translation that preserves intuitionistic validity and reduces formulas to a normal form.

@@

The interest in intuitionistic logic has lead to the development of a number of automated theorem proving systems for propositional as well as for predicate logic and a collection of benchmark problems (see e.g. the
ILTP library website~\cite{iltp}).
However, progress in automated theorem proving for intuitionistic logic has been slow, whereas reasoners for classical logic have made tremendous progress, see e.g. the TPTP~\cite{casc} and SAT~\cite{satc} competitions.
%, e.g., the first-order provers~\cite{kovacs2013first, schulz2002brainiac, korovin2008iprover}
%Comparing benchmarks for classical and intuitionistic automated theorem proving, e.g. TPTP~\cite{casc} and , we see that classical provers currently have a far better success rate. Furthermore all the most popular first-order utilize classical logic.
This difference can be partially explained by fundamental differences between the logics.
First of all, determining intuitionistic validity is computationally harder, i.e. in the propositional case intuitionistic validity is \verb+PSPACE+-complete~\cite{statman1979intuitionistic} whereas classical validity is \verb+coNP+-complete~\cite{cook1971complexity}.
A further advantage of classical logic is the existence of calculi that are particularly suitable for automation such as superposition~\cite{bachmair2001resolution}, which rely on the existence of convenient normal forms such as CNF, and the duality between validity and satisfiability, i.e. to show the validity of a formula it suffices to show the unsatisfiability of the negated formula.
While some (albeit more complex) normal forms also exist  for intuitionistic logic, crucially the duality between validity and satisfiability of the negation does not hold.
Therefore most dedicated intuitionistic theorem provers~\cite{mclaughlin2009efficient, tammet1996resolution} use the naive inverse method, i.e. directly search for a cut-free proof by applying the rules from some proof calculus inversely, enhanced by search strategies such as focussing and polarization. This approach generally leads to a much more complex search and is therefore difficult to implement efficiently.
Finally, we add that in contrast to intuitionistic provers a tremendous amount of work has been put into optimizing provers for classical logic, in particular for the propositional case, i.e. SAT-solvers.

@@

\textbf{Related Work.}
Several approaches to obtain constructive content from classical proofs have been studied in the literature.
An important direction is to examine under which circumstances classical validity entails intuitionistic validity and therefore constructiveness.
This entailment has most famously been established for $\Pi_2$-formulas in Peano Arithmetic~\cite{friedman1978classically}, but also other settings have been studied~\cite{schwichtenberg}.
Another important direction is to equip classical proofs with operational semantics~\cite{Control1, Parigot1}, which however does not permit in general the realizability of at least $\exists$ and $\vee$.

@@

\begin{theorem}\label{thm:Tseytin}
For every propositional formula $\varphi$ there effectively are an atom $P$ and a set of $\mathcal S$ of formulas (over an extended set of propositions) of one of the forms\\
$\text{ \hspace{1cm}} A\to (B\wedge C), (A\wedge B)\to C, A\to (B\vee C), (A\vee B)\to C, (A\to B)\to C,$\\
for atomic $A, B, C$, such that $\varphi$ is intuitionistically valid if and only if $\bigwedge\mathcal S\to P$ is intuitionistically valid. The time complexity of the transformation is linear in the input size.
\end{theorem}

@@

It is known that a polynomial time translation between QBF and IPC must exist, since both problems are PSPACE complete~\cite{garey1979computers, statman1979intuitionistic}.
However, giving an explicit translation from IPC to QBF still promises to be useful, allowing to leverage the progress in QBF solving.
The translation from the last section will serve as starting point for developing the translation in this section.
For lack of space, we do not define the semantics of QBF in this paper and refer the reader to standard sources e.g.~\cite{series/faia/2009-185}.

@@

\bibliography{lipics-v2021-EIICL}

>>>>

\documentclass[runningheads]{llncs}
\usepackage{amsmath, amsfonts}
\usepackage{wrapfig}
\usepackage{tikz-qtree}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{arrows,shapes,quotes}
% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%

%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
	%
	\title{Embedding Intuitionistic into Classical Logic}
	%
	%\titlerunning{Abbreviated paper title}
	% If the paper title is too long for the running head, you can set
	% an abbreviated paper title here
	%
	\author{Alexander Pluska\inst{1}\orcidID{0000-0002-7709-3335} \and
		Florian Zuleger\inst{2,3}\orcidID{0000-0003-1468-8398}}
	%
	\authorrunning{A. Pluska and F. Zuleger}
	% First names are abbreviated in the running head.
	% If there are more than two authors, 'et al.' is used.
	%
	\institute{TU Wien, Vienna, Austria}
	%
	\maketitle
\begin{abstract}
The famous double negation translation~\cite{glivenko1929quelques,godel1933intuitionistischen} establishes an embedding from classical into intuitionistic logic. Curiously the reverse direction has not been covered in literature. We present an effective embedding from intuitionistic into classical logic, both in the propositional and first-order case, as well as an effective embedding of intuitionistic propositional logic into quantified boolean formulas.
Furthermore, in the propositional case we establish a parameter that bounds the proof complexity of intuitionistic formulas. As part of our construction, we also present a Tseytin-like translation that preserves intuitionistic validity and reduces formulas to a normal form.

@@

The interest in intuitionistic logic has lead to the development of a number of automated theorem proving systems for propositional as well as for predicate logic and a collection of benchmark problems (see e.g. the
ILTP library website~\cite{iltp}).
However, progress in automated theorem proving for intuitionistic logic has been slow, whereas reasoners for classical logic have made tremendous progress, see e.g. the TPTP~\cite{casc} and SAT~\cite{satc} competitions.
%, e.g., the first-order provers~\cite{kovacs2013first, schulz2002brainiac, korovin2008iprover}
%Comparing benchmarks for classical and intuitionistic automated theorem proving, e.g. TPTP~\cite{casc} and , we see that classical provers currently have a far better success rate. Furthermore all the most popular first-order utilize classical logic.
This difference can be partially explained by fundamental differences between the logics.
First of all, determining intuitionistic validity is computationally harder, i.e. in the propositional case intuitionistic validity is \verb+PSPACE+-complete~\cite{statman1979intuitionistic} whereas classical validity is \verb+coNP+-complete~\cite{cook1971complexity}.
A further advantage of classical logic is the existence of calculi that are particularly suitable for automation such as superposition~\cite{bachmair2001resolution}, which rely on the existence of convenient normal forms such as CNF, and the duality between validity and satisfiability, i.e. to show the validity of a formula it suffices to show the unsatisfiability of the negated formula.
While some (albeit more complex) normal forms also exist  for intuitionistic logic, crucially the duality between validity and satisfiability of the negation does not hold.
Therefore most dedicated intuitionistic theorem provers~\cite{mclaughlin2009efficient,tammet1996resolution} use the naive inverse method, i.e. directly search for a cut-free proof by applying the rules from some proof calculus inversely, enhanced by search strategies such as focussing and polarization. This approach generally leads to a much more complex search and is therefore difficult to implement efficiently.
Finally, we add that in contrast to intuitionistic provers a tremendous amount of work has been put into optimizing provers for classical logic, in particular for the propositional case, i.e. SAT-solvers.

@@

\textbf{Related Work.}
Several approaches to obtain constructive content from classical proofs have been studied in the literature.
An important direction is to examine under which circumstances classical validity entails intuitionistic validity and therefore constructiveness.
This entailment has most famously been established for $\Pi_2$-formulas in Peano Arithmetic~\cite{friedman1978classically}, but also other settings have been studied~\cite{schwichtenberg}.
Another important direction is to equip classical proofs with operational semantics~\cite{Control1,Parigot1}, which however does not permit in general the realizability of at least $\exists$ and $\vee$.

@@

\begin{theorem}\label{thm:Tseytin}
For every propositional formula $\varphi$ there effectively are an atom $P$ and a set of $\mathcal S$ of formulas (over an extended set of propositions) of one of the forms\\
$\hspace{1cm} A\to (B\wedge C), (A\wedge B)\to C, A\to (B\vee C), (A\vee B)\to C, (A\to B)\to C,$\\
for atomic $A, B, C$, such that $\varphi$ is intuitionistically valid if and only if $\bigwedge\mathcal S\to P$ is intuitionistically valid. The time complexity of the transformation is linear in the input size.
\end{theorem}

@@

It is known that a polynomial time translation between QBF and IPC must exist, since both problems are PSPACE complete~\cite{garey1979computers,statman1979intuitionistic}.
However, giving an explicit translation from IPC to QBF still promises to be useful, allowing to leverage the progress in QBF solving.
The translation from the last section will serve as starting point for developing the translation in this section.
For lack of space, we do not define the semantics of QBF in this paper and refer the reader to standard sources e.g.~\cite{series/faia/2009-185}.

@@

\bibliographystyle{splncs04}
\bibliography{bibliography}
